 

### 2020-11-13 doc/manuscript/Dow_manuscript.Rmd
add tree-rings, format, misc

---
title: ''
output:
  word_document:
    fig_caption: yes
bibliography: references.bib
csl: global-change-biology.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

#Figure numbers
schematic=1 #illustrates parameters
growth_curves_all = 2  
    rel_growth_SCBI = "2a" 
    rel_growth_HF = "2b" 
    cum_growth_SCBI = "2c" 
    cum_growth_HF = "2d" 
climwin = 3
pheno_Tsensitivity = 4
DOYtiming = 5
DOYtiming_SCBI = "5a"
DOYtiming_Harvard = "5b"

@@

@@

### Abstract

@@

### Introduction

#### (1. forests are critical for climate change regulation, so we need to understand better how they're responding to climate change)
As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented in response, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid. The absence of human intervention increases our reliance on natural systems to soften the blow of climate change. Plant life, especially forests, make up a large portion of these natural systems, offering many valuable ecosystem services. For one, they house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO2 from the atmosphere and storing it in the trees or soil. Forests sequester about one-third of all anthropogenic CO~2~ and store ~45% of all terrestrial carbon [@bonan_forests_2008]. However, their future under the current era of rapid change is uncertain [@friedlingstein_climatecarbon_2006].

@@

#### (2. spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
In recent decades, an increase in average spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). 
Shifts in leaf phenology timing could be accompanied by changes in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;). 
The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations*. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011). The timing of this spring warming also has a significant effect on leaf phenology response (Crabbe et al, 2016).

@@

#### (3. However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )

@@

#### (4. We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
~~A recent publication by D'Orangeville et al () found that~~ Leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous; **D'Orangeville et al. YY**). For example, ~~they observed that~~ bud burst occurs around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.

@@

#### (5. Here, we...)
Here, we characterize how early spring temperatures affect stem growth phenology and rates of temperate deciduous trees within **two/three** forest dynamics plots in the Eastern USA. Using data from dendrometer bands measured throughout the growing season, we fit a growth model  to the time series of individual trees to determine the day of year (DOY) where 25, 50, and 75% annual growth was achieved; maximum growth rates and the DOY on which they occurred; total annual growth; and peak growing season length (75%-25% DOY) [Fig. `r schematic`; @mcmahon_general_2015]. We test the hypotheses that (1) warmer early springs result in earlier stem growth and a period of growth, but (2) maximum growth rates are independent of spring temperatures, and as a result (3) total annual growth increases in response to warmer spring temperatures (Table `r hypothesis_table`).

@@

### Materials and Methods

#### Study sites and data

@@

*(Do we want a table of species here?? Columns would be Species, xylem porosity, n trees, n tree-years (after cleaning). Details will be in an SI table)*

@@

@@

#### Fitting growth curves

@@

#### Determining optimal climate windows

@@

#### Hypothesis testing

@@

### Results

@@

### Discussion

@@

### (Content to incorporate)

@@

### Acknowledgements
ForestGEO

### Author Contributions

### Supporting Information

@@

### References

>>>>

---
title: ''
output:
  bookdown::word_document2: 
    reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    fig_caption: yes
bibliography: references.bib
csl: global-change-biology.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

#Figure numbers
schematic=1 #illustrates parameters
growth_curves_all = 2  
    rel_growth_SCBI = "2a" 
    rel_growth_HF = "2b" 
    cum_growth_SCBI = "2c" 
    cum_growth_HF = "2d" 
climwin = 3
pheno_Tsensitivity = 4
DOYtiming = 5
DOYtiming_SCBI = "5a"
DOYtiming_Harvard = "5b"
tree_rings = 6

@@

tree_rings_caption <- "Figure 6. Climate sensitiivity of annual growth from tree-rings. THIS IS JUST A PLACEHOLDER TO SHOW THE RESULTS--WE'LL WANT TO RE-MAKE FIGURE."

@@

# Abstract

@@

# Introduction

## (1. forests are critical for climate change regulation, so we need to understand better how they're responding to climate change)
As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented in response, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid. The absence of human intervention increases our reliance on natural systems to soften the blow of climate change. Plant life, especially forests, make up a large portion of these natural systems, offering many valuable ecosystem services. For one, they house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO2 from the atmosphere and storing it in the trees or soil. Forests sequester about one-third of all anthropogenic CO~2~ and store ~45% of all terrestrial carbon [@bonan_forests_2008]. However, their future under the current era of rapid change is uncertain [@friedlingstein_climatecarbon_2006].

@@

## (2. spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
In recent decades, an increase in average spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). 
Shifts in leaf phenology timing could be accompanied by changes in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;). 
The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations*. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011). The timing of this spring warming also has a significant effect on leaf phenology response (Crabbe et al, 2016).

@@

## (3. However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )

@@

## (4. We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
~~A recent publication by D'Orangeville et al () found that~~ Leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous; **D'Orangeville et al. YY**). For example, ~~they observed that~~ bud burst occurs around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.

@@

## (5. Here, we...)
Here, we characterize how early spring temperatures affect stem growth phenology and rates of temperate deciduous trees within two forest dynamics plots in the Eastern USA. Using data from dendrometer bands measured throughout the growing season, we fit a growth model  to the time series of individual trees to determine the day of year (DOY) where 25, 50, and 75% annual growth was achieved; maximum growth rates and the DOY on which they occurred; total annual growth; and peak growing season length (75%-25% DOY) [Fig. `r schematic`; @mcmahon_general_2015]. We test the hypotheses that (1) warmer early springs result in earlier stem growth and a period of growth, but (2) maximum growth rates are independent of spring temperatures, and as a result (3) total annual growth increases in response to warmer spring temperatures (Table `r hypothesis_table`).

@@

# Materials and Methods

## Study sites and data

@@

\newpage
*(Do we want a table of species here?? Columns would be Species, xylem porosity, n trees, n tree-years (after cleaning). Details will be in an SI table)*

@@

(*tree rings, if included-- KAT can write this*)

@@

## Fitting growth curves

@@

## Determining optimal climate windows

@@

## Hypothesis testing

@@

# Results

@@

*Climate sensitivity of growth*

```{r tree_rings, out.width="100%", echo=FALSE, fig.align="center", fig.cap= tree_rings_caption}
# Created in RScripts/additional_figures.R
knitr::include_graphics("tables_figures/tree_rings.tif")
```

# Discussion

@@

# (Content to incorporate)

@@

# Acknowledgements
people: (everyone who helped collect data), Jess Shue (data), Neil Pederson (helpful discussion) 
Funding: ForestGEO

# Author Contributions

# Supporting Information

@@

# References

### 2020-11-13 doc/manuscript/tables_figures/tables.Rmd
add tree-rings, format, misc

kable(hypothesis_table, col.names = c("Hypotheses and Specific Predictions", "RP", "DP","RP", "DP"), caption = "Table 1. Hypothesis table", format = "latex", booktabs = TRUE,  escape = F)%>%
  #column_spec(1, width = "4cm") %>%
  add_header_above(c(" ", "SCBI" = 2, "Harvard"=2 )) %>%
  pack_rows("1. Warmer early springs result in earlier stem growth and longer growing seasons", 1, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("2. Maximum growth rates are independent of early spring temperatures.", 6, 6, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("3. Annual stem growth responds positively to warmer spring temperatures.", 7, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
```

>>>>

kable(hypothesis_table, col.names = c("Hypotheses and Specific Predictions", "RP", "DP","RP", "DP"), caption = "Table 1. Hypothesis table", format = "latex", booktabs = TRUE,  escape = F)%>%
  #column_spec(1, width = "4cm") %>%
  add_header_above(c(" ", "SCBI" = 2, "Harvard"=2 )) %>%
  pack_rows("1. Warmer early springs result in earlier stem growth and longer growing seasons", 1, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("2. Maximum growth rates are independent of early spring temperatures.", 6, 6, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("3. Annual stem growth responds positively to warmer spring temperatures.", 7, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
```
